Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-03-31 review/answer_to_review.md

I was not aware of this work and added a paragraph in the introduction to reference it. It is indeed relevant.

> In the paper, it’s very high the amount of data, for which the archaeological information regarding the type of burial and funerary rite is lacking. At table 1, page 4, we can clearly detect that for 60% of the dates associated to cremation burials we have no information regarding the type of funerary structure. For 37% of the inhumations, no information about the funerary structure is available. Moreover, for 32% of the dataset (559 radiocarbon dates out of 1701) we do not have any kind of information regarding neither the funerary rite nor the burial structure. This lack of data obliges the author to consider carefully the results reached in the paper. It’s relevant to highlight all the biases of the prosed research and how the conclusion can be misleading due to the quality of available data. In particular, when the statistical analyses carried out on data aim to explore in the details the cultural distance among the different regions.

This is true and I tried to emphasise the limitations of the data in the discussion. The challenges are not confined on the amount of data, but also concern its meaning and origin. The supplementary remarks cover these problems in detail.

I now modified the Figures 3&4 as well as the Table 1 to provide an overview over the actual amount of data available. I have also added several short comments/sentences in various sections of the paper to better address the small sample size issue.

@@

## Comments to the Author:

> I had the chance to read the manuscript entitled “Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data” authored by C. Schimd. This is an interesting paper that combines empirical analysis of Bronze age burial sites across Europe with a computer simulation of intergroup cultural transmission. The manuscript is well written, I enjoyed reading it, and think that the core findings are definitely of interest for archaeologist working on Bronze age Europe and for those interested in the application of cultural evolutionary framework. I am thus keen to recommend its publication, although I think several issues needs to be addressed before this.

> Firstly, my impression is that the author is trying to squeeze perhaps a bit too much in a single paper. I felt this was particularly the case of the computer simulation model which is almost thrown in in the mix, without much discussion on model parameters (see below). I would be tempted to suggest removing this entirely and include this into another paper.

After careful deliberation, I decided to follow this advice. The simulation (1) does not add information that would be absolutely essential for the core topics, (2) reduces the overall comprehensibility due to the terminology that has to be introduced to describe it and (3) cannot be adequately presented and discussed here without overstraining the initial scope of the paper.

I removed the simulation here. A future paper will present it together with other simulation approaches.

@@

> Secondly, I feel like the empirical analyses have too much preprocessing. The raw data provides spatial and temporal coordinates of individual burials, yet these are combined spatially and chronologically first and analysed as aggregates. As a consequence: 1) there is a substantial loss of spatial information; 2) sampling error is effectively disregarded (i.e. mantel tests are based on proportion values that are calculated for regions with very different samples sizes) and strictly speaking the samples are also non-independent (i.e. a large site with multiple burials can strongly bias the proportions of a particular region); and 3) there is an increased risk of ecological fallacy/Simpson’s paradox. I take this aggregation provides a direct mapping to existing literature, but I am not convinced that this is worth sacrificing so much.

The decision to create discrete spatial area classes is necessary to establish subsets of dates that can be used for time series calculation. I'm not aware of a methodological approach where time series could be constructed without some sort of aggregation. The equal area size approach applied here for the region definition causes their size to be very big due to the currently overall small amount of data. I expect this to change in the future and the analysis can be redone then with more samples and much smaller areas.

So: The criticism listed here is valid and I added it in the *Discussion* section. I considered a complete revision of the time series construction methodology, but since I (1) am still convinced that it is scientifically sound and reproducible, and (2) do not know a fundamentally better approach that avoids these risks completely, I suggest publishing as it is for the community of practice to discuss and improve.

@@

> Thirdly, the author implies that the scripts used for his analysis are supplied but there are no links on the article to any repository and I did not find any electronic supplementary material. As the author is well aware, we need to do our best to make our work fully reproducible.

This is a result of miscommunication with the special issue editors. I added links to the relevant repositories now.

@@

> Page 4, Columns 1-2, Lines 58-60; 1-8. This is a major issue. Perhaps it is worth bootstrap confidence interval for these proportions? This way both sample size and sample independence could be tackled. Also, please provide the sample size for each region.

I'm very interested in methods to assign a measure of confidence to the results for individual regions and times. Figure 3&4 are deliberately constructed to juxtapose the relative proportion with the amount of graves available. I understand this as an object of future research. I reworked Table 1 and Figure 3&4 for a better presentation of the sample size per region.

> Page 4, Column 2, Lines 43-56. I think it is worth computing partial mantel tests so that the correlation between burial type and construction can take into account geography.

A good idea! Done. I split Table 2 to present the results, added them to the new version of Figure 6 and addressed them in the text.

### 2019-05-23 review/answer_to_review_2.md

# Associate Editor

Comments to the Author:

> Dear author, 
> I agree with the comments of the reviewer and the remaining points have to be addressed before we can accept this paper.  As some points imply adding new analyses with different tools, it may not be reasonable to  do it in the scope of this paper. But at least a stronger and clearer emphasis on the flaws and weaknesses of the statistical analysis should be done.  The alternatives approach proposed by the reviewer should be at the very least mentioned. It is not a problem if the result are, on the overall, not highly significant __as far as__ it's clear in the text. The approach and the methods are already a a valuable contribution to our special issue. I will look forward to receiving a revised manuscript from you.

TODO

# Reviewer: 1

Comments to the Author:

> The methods adopted in the paper are very innovative and have been rarely implemented to analise large 14C-dates datasets. I am strongly conviced of the quality of the manuscript, that has improved in this second version.

> Only a small remark, the publication year of the paper "Capuzzo G, Boaretto E, Barceló JA" is 2014.

TODO

# Reviewer: 2

Comments to the Author:

> I read the revised version of the manuscript and I think the authors have addressed most of the issues I raised, and acknowledged others in the text. I am still unconvinced by the creation of the artificial subsets and think that perhaps using temporal distance matrix (see for example Shennan et al 2015 on Evolution and Human Behaviour) might have been a solution, but I also see the point that such approach would not directly explore time-series. Ultimately this is an instance of the so-called MAUP (Modifiable Areal Unit Problem) in geography, and perhaps there are some more nuanced approaches out there. I think for the purpose of this paper acknowledging the problem is more than sufficient.  I also still have some issues with how sample size is effectively being ignored once proportions are computed, and I think this should be emphasised more strongly as I expect not everyone would be reading beyond the conclusion section.

TODO

> Please find below some further minor comments on the revised text. Overall, I think this is a great piece of work and I will look forward to follow-ups, including the simulation work that was excluded from this manuscript.

> Page 2, Column 1, Line 29. ‘simplify survival’ → increase?

TODO

> Page 2, Column 1, Line 42. ‘seemlingly random’  This is one of those tricky phrasing that has different meaning depending on the readership. ‘Random’ to me almost equals to unbiased transmission with drift, and stating that it is ‘seemlingly’ reads as if in reality this is not the case. This is perfectly fine, but it is slightly at odds with the definition of ‘style’ and the implicit assumption of neutrality. I would consider rephrasing, just to avoid potential misunderstandings.

TODO

> Page 2, Column 2, Line 41. the definition of drift here is more focused on its consequence. The fixation scenario presented applies only in regimes with no mutation/innovation. Perhaps specify that with no innovation a single trait will dominate by chance alone due to chance alone?

TODO

> Page 3, Column 1, Lines 2-6. Great!

> Page 3, Column 2, Line 59. ‘the region are circular’ Why this shape if it causes overlap? Why not hexagons?

TODO

> Page 5, Figure 3. Just an idea for a follow-up study. Given these are effectively adoption curves, one could fit a logistic model to assess differences in the rate of adoption and estimates on the timing of earliest uptake. It would be also very interesting to determine whether indeed there is a long-tail in southern Germany and what this means (cf. Henrich 2001 on American Anthropologist, but see also recent debates on adoption curves, e.g. https://www.biorxiv.org/content/biorxiv/early/2017/11/16/159038.full.pdf). Food for thoughts.

TODO

> Page 5, Column 1, Lines 56-. Please specify the R package used for the Mantel and Partial Mantel Tests.

TODO

> Page 5. I still think a permutation test would be more robust in this case. One could compare the observed correlation between two pairs of proportion time series against a simulation set where each iteration is the measured correlation between the proportion time-series of two artificial sets generated by randomly permuting the location label of each radiocarbon date. Pairs with a smaller number of dates will generate a wider range of correlation values and will be harder to test. One could adopt a more sophisticated solution by permuting groups of dates that occurred together in order to handle sample interdependence. Might be worth considering doing this for a follow-up study.

TODO

> Page 6, Column 2, Lines 46-49. ‘If one believes the sample, then a sudden, radical change from burial in flat graves to hill graves took place in Southern Germany at the beginning of the Middle Bronze Age.’  The sample size here is 8 dates. I appreciate the cautionary ‘if one believes the sample’ but this is a bit of a stretch to say the least, and readers that are numerically less literate might be misled. The same applies to other contexts with extremely small sample sizes and I think adding confidence intervals to those proportions would help.

TODO

> Table 3 & Figure 6. The figure nicely summarises the trend detailed on Table 3, but does not distinguish instances where the correlation were identified as signficant vs those that were not. Indeed in no cases the analysis shows a signficant correlation under the Partial Mantel test, whilst only few chronological blocks are showing signficant correlations. I suggest highlighting this by perhaps changing the the point symbology. It is also worth noting that if anything there is a bias towards type I error given: a) that we are dealing with multiple testing; and b) these analyses do not take into account sample sizes which are often very small. P-values are not the core issue here, but rather the fact that some way to at least visually convey this uncertainty is much needed.

TODO

> Page 11, Column 1, line 54. Yes, but none of the negative correlation are statistically signficant!

TODO

### 2019-05-23 article/02_Materials_and_Methods.Rmd
full stop

The code for this paper is written in the statistical programming language R [@rcoreteamLanguageEnvironmentStatistical2018] and fully reproducible [@marwickPackagingData2018]. It is available in a public Repository on GitHub (\url{https://github.com/nevrome/cultrans.bronzeageburials.article2019}) and with a permanent digital object identifier (DOI) on the Open Science Framework Platform (\url{http://dx.doi.org/10.17605/OSF.IO/B6NP2}) A more in-depth description and discussion of the data preparation steps can be found in the master thesis this article is based on [@schmidComputerbasiertesCulturalEvolution2018].

>>>>

The code for this paper is written in the statistical programming language R [@rcoreteamLanguageEnvironmentStatistical2018] and fully reproducible [@marwickPackagingData2018]. It is available in a public Repository on GitHub (\url{https://github.com/nevrome/cultrans.bronzeageburials.article2019}) and with a permanent digital object identifier (DOI) on the Open Science Framework platform (\url{http://dx.doi.org/10.17605/OSF.IO/B6NP2}). A more in-depth description and discussion of the data preparation steps can be found in the master thesis this article is based on [@schmidComputerbasiertesCulturalEvolution2018].
